---
title: "Cristina Lazcano and Eric Boyd - strawberry endophyte and soil metagenomics 16s time course"
author: "Sam Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, echo=FALSE, results="hide"}

# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("dada2", "biomformat", "phyloseq", "pheatmap", "tidyr")

library('knitr')
knitr::opts_chunk$set(echo = TRUE)

library(dada2); packageVersion("dada2")
library(biomformat); packageVersion("biomformat")
library(phyloseq); packageVersion("phyloseq")
library(ggplot2); packageVersion("ggplot2")
#library(DECIPHER); packageVersion("DECIPHER")
#library(phangorn); packageVersion("phangorn")
library(pheatmap); packageVersion("pheatmap")
library(tidyr)
library(kableExtra)
library(Biostrings)

options(stringsAsFactors=F)

wd = '/share/biocore/projects/Lazcano_Cristina_UCD/2020-06-22-Cristina_Eric-LandWaterSoil/16s.analysis/02-DADA2-analysis'
knitr::opts_knit$set(root.dir = wd)
setwd(wd)

```

## Reading data etc

```{r setup_filter}

#
path <- "../01-HTS_Preproc/" # CHANGE ME to the directory containing the fastq files after unzipping.
#path <- "../02-Overlap" # CHANGE ME to the directory containing the fastq files after unzipping.
list.files(path)

# Read in "SE" files:
# Note that all of the NTCs were not sequenced.
fnFs.SE = sort(list.files(path, pattern="*_SE.fastq", full.names=T))

#sample.names <- paste("s", sapply(strsplit(basename(fnFs.SE), "_SE"), `[`, 1), sep='')
sample.names <-  sapply(strsplit(basename(fnFs.SE), "_SE"), `[`, 1)

#pdf(file="QC_plots.pdf", w=11, h=8)
# # Examine quality profile plots:
png("fnFs_quality.png", width=2000, height=1000)
plotQualityProfile(fnFs.SE[1:11])
dev.off()

### Note that filtering and trimming should not be necessary if reads are filtered for any
### containing an "N"
#Perform filtering and trimming
#Assign the filenames for the filtered fastq.gz files.
# filt_path <- file.path("01-filtered") # Place filtered files in filtered/ subdirectory
# filtFs <- file.path(filt_path, paste0(sample.names, "_F_filt.fastq.gz"))

#Filter the forward and reverse reads
# out <- filterAndTrim(fwd=fnFs.SE, filt=filtFs, truncLen=0, minLen=350,
#               maxN=0, maxEE=Inf, truncQ=0, rm.phix=F, 
#               trimLeft=0, trimRight=0, minQ=0, rm.lowcomplex=0, orient.fwd = NULL,
#               compress=TRUE, multithread=TRUE, verbose=TRUE)


# This filter is pretty relaxed, we keep almost all reads.  
# out <- as.data.frame(out)
# out$Pct = 100 * out[,"reads.out"]/out[,"reads.in"]
# out

```

```{r errors}
# Learn error rate with forward reads:
errF <- learnErrors(fnFs.SE, multithread=TRUE)

png("errors.png", width=2500, height=1500)
    plotErrors(errF, nominalQ=TRUE)
dev.off()

```

```{r dada2}
dadaUs = dada(fnFs.SE, err=errF, multithread=TRUE, pool=T, OMEGA_C=0)  # OMEGA_C=0 ?
#saveRDS(dadaUs, "dadaUs.RDS")  # This is really large..
gc()
#saveRDS(dadaUs, file="dadaUS.rds") # This file is huge and takes forever to save, don't bother.

```

```{r dada_summaries}
#dadaUs.OMC = dada(filtFs, err=errF, multithread=TRUE, pool=T, OMEGA_C=0)  # OMEGA_C=0 ?

#table(dadaUs$sCerus_1_F_filt.fastq.gz$map, useNA = "always")

#dadaUs[[1]]
#length(dadaUs)
#sum(dadaUs[[1]]$clustering$abundance)

#length(dadaUs[[1]]$map)
#sum(getUniques(dadaUs[[1]]))
#sum(is.na(dadaUs[[1]]$map))

#head(getSequences(dadaUs[[1]]))
#head(getUniques(dadaUs[[1]]))

# Construct sequence table:
seqtab <- makeSequenceTable(dadaUs)
rownames(seqtab) = sample.names
save(seqtab, file="seqtab.RData")
dim(seqtab)


# Inspect distribution of sequence lengths:
table(nchar(getSequences(seqtab)))

#  315   341   343   358   359   365   376   377   379   380   382   383   384 
#     4     2     1     1     1     1     4     3     1     3     3     1    26 
#   385   386   387   389   394   395   396   397   398   399   400   401   402 
#     5     3    10     1     1     1     1     1     5     7    44   316  9922 
#   403   404   405   406   407   408   409   410   411   412   413   414   415 
#  3224  1834  1118   391  4088   418   463   203   345   119   321   134    88 
#   416   417   418   419   420   421   422   423   424   425   426   427   428 
#   415   172   192  1165   519  1368  3560   448   798   532  4866 10670  3883 
#   429   430   431   432   433   435   436   437   438   439   440   441   442 
#   427    59     9     1     1    18    10    66    26    22     3     1     3 
#   443   444   445   446   447   448   450 
#     2     2    10     4     3     5     6 


```

DADA2 recommends a chimera removal step. From what I can tell, this step takes ASVs with very high read count support and then looks for low-support ASVs that have perfect matches to the high-support ASVs over part of their length. If a low-support ASV appears to be a combination of two low-support ASVs, the low support ASV is removed from the dataset.

```{r}

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
save(seqtab.nochim, file="seqtab.nochim.RData")

# Identified 3257 bimeras out of 7538 input sequences.

# Try removal with more stringent n-fold parameter
# seqtab.nochim = removeBimeraDenovo(seqtab, method="consensus", minFoldParentOverAbundance=8, multithread=TRUE, verbose=TRUE)
# Identified 4813 bimeras out of 8270 input sequences
## TODO: Investigate removeBimeraDenovo further. It seems to be removing a lot of things that are exactly/almost exactly the same
# length as the rest of the amplicons in the study, but that seems surprising given that they are supposedly chimeric sequences.

gc()
dim(seqtab.nochim)
# [1]   213 45383


# What faction of total bases is retained?
sum(seqtab.nochim)/sum(seqtab)
# [1] 0.9503154


# Sequence lengths before nochim:
table(nchar(getSequences(seqtab)))


#   315   341   343   358   359   365   376   377   379   380   382   383   384 
#     4     2     1     1     1     1     4     3     1     3     3     1    26 
#   385   386   387   389   394   395   396   397   398   399   400   401   402 
#     5     3    10     1     1     1     1     1     5     7    44   316  9922 
#   403   404   405   406   407   408   409   410   411   412   413   414   415 
#  3224  1834  1118   391  4088   418   463   203   345   119   321   134    88 
#   416   417   418   419   420   421   422   423   424   425   426   427   428 
#   415   172   192  1165   519  1368  3560   448   798   532  4866 10670  3883 
#   429   430   431   432   433   435   436   437   438   439   440   441   442 
#   427    59     9     1     1    18    10    66    26    22     3     1     3 
#   443   444   445   446   447   448   450 
#     2     2    10     4     3     5     6 


# Look at distribution of fragment lengths discarded by the chimera removal step.
table(nchar(setdiff(getSequences(seqtab),getSequences(seqtab.nochim))))

#  341  365  380  383  395  398  400  401  402  403  404  405  406  407  408  409 
#    1    1    2    1    1    1   10   12 2121  162  265   55   39 1594   42   38 
#  410  411  412  413  414  415  416  417  418  419  420  421  422  423  424  425 
#    3   40   18    8   12   14  109   19   16  119   57   41  278   21   19   53 
#  426  427  428  429  431  436  437  442  443  445 
#  273 1373  158   14    1    1    1    1    1    2 


```

#### Were the lengths of discarded chimeras the same as the original sequences?

```{r PlotChimeraLengthDistribution}

par(mfrow=c(2,1))
barplot(table(nchar(getSequences(seqtab))),  
    main="Sequence length distribution of all ASVs")
barplot(table(nchar(setdiff(getSequences(seqtab),getSequences(seqtab.nochim)))), 
    main="Sequence length distribution of discarded bimeras")


```


```{r}

#Track reads through the pipeline
getN <- function(x) sum(getUniques(x))

track = data.frame(input = sapply(dadaUs, getN),
                   denoised = rowSums(seqtab),
                   nochim = rowSums(seqtab.nochim),
                   ASVs = rowSums(seqtab>0),
                   ASVs.nochim = rowSums(seqtab.nochim>0))

#track <- cbind(out, sapply(dadaUs, getN), rowSums(seqtab), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
#colnames(track) <- c("input", "filtered", "pctfiltered", "denoised", "tabled", "nonchim")
rownames(track) <- sample.names
track

write.table(data.frame(sample=rownames(track), track), file="read_tracking.tsv", sep='\t', row.names=F)
```


```{r heatmap}
pdf(file='Top20_AVS_read_count_heatmap.pdf', width=8, height=16)
df = seqtab.nochim
seqs = colnames(df)
colnames(df) = paste0("ASV", 1:ncol(df))
names(seqs) = colnames(df)
df = df[grep('Blank', rownames(df), invert=T), ]
df = df[grep('blank', rownames(df), invert=T), ]
df = df[grep('Zymo', rownames(df), invert=T), ]
df = df[,order(colSums(df), decreasing=T) ]
pheatmap(df[,1:20], main='Read count for top 20 ASVs amplicons', cluster_cols=F)
dev.off()
```


```{r silva_taxonomy}
# Assign taxonomy to nochim data: 
taxa.silva.nochim = assignTaxonomy(seqtab.nochim, "./SILVA/silva_nr_v138_train_set_strawberry.fa.gz", multithread = T, minBoot=50)
taxa.silva.nochim = addSpecies(taxa.silva.nochim, "./SILVA/silva_species_assignment_v138_strawberry.fa.gz")
save(taxa.silva.nochim, file="taxa.silva.nochim.RData")
gc()
colSums(!is.na(taxa.silva.nochim))

# Without strawberry in reference:
# Kingdom  Phylum   Class   Order  Family   Genus Species 
#   45371   44616   43877   40283   33991   21505     945 

# With strawberry in reference:
# Kingdom  Phylum   Class   Order  Family   Genus Species 
#   45364   44637   43885   40267   34067   21619     948 

# With strawberry and second plant in reference:
# Kingdom  Phylum   Class   Order  Family   Genus Species 
#   45361   44595   43865   40286   34070   21657     947 

#
#

```{r compare_nochim}
### Not run because of the massive number of ASVs generated.

# Compare assignments at the Genus level:
# library(phyloseq)
# ASVs.nochim = DNAStringSet(colnames(seqtab.nochim))
# names(ASVs.nochim) = paste0("ASV", 1:ncol(seqtab.nochim))
# tmp.seqtab = seqtab.nochim
# colnames(tmp.seqtab) = names(ASVs.nochim)
# tmp.taxa = taxa.silva.nochim
# rownames(tmp.taxa) = names(ASVs.nochim)
# ps.nochim = phyloseq(otu_table(tmp.seqtab, taxa_are_rows=FALSE),
#              tax_table(tmp.taxa),
#              refseq(ASVs.nochim))

# ps.nochim.family = tax_glom(ps.nochim, "Family")

# # Get total ASVs per Family:
# nochim.df = data.frame(Family = tax_table(ps.nochim.family)[,5], ASVcount = colSums(otu_table(ps.nochim.family)))


# ASVs.all = DNAStringSet(colnames(seqtab))
# names(ASVs.all) = paste0("ASV", 1:ncol(seqtab))
# tmp.seqtab = seqtab
# colnames(tmp.seqtab) = names(ASVs.all)
# tmp.taxa = taxa.silva.all
# rownames(tmp.taxa) = names(ASVs.all)
# ps.all = phyloseq(otu_table(tmp.seqtab, taxa_are_rows=FALSE),
#              tax_table(tmp.taxa),
#              refseq(ASVs.all))
# ps.all.family = tax_glom(ps.all, "Family")
# all.df = data.frame(Family = tax_table(ps.all.family)[,5], ASVcount = colSums(otu_table(ps.all.family)))

# # Compare:
# all.family = unique(c(nochim.df$Family, all.df$Family))
# combined.df = data.frame(Family = all.family, 
#                          nochim = nochim.df$ASVcount[match(all.family, nochim.df$Family)],
#                          all = all.df$ASVcount[match(all.family, all.df$Family)])

# write.table(combined.df, "compare_nochim.tsv", row.names=F, sep='\t')

# kable(combined.df) %>%
#   kable_styling("striped", full_width = F) %>%
#   row_spec(0, angle = 0)

#ps.nochim.genus = tax_glom(ps.nochim, "Genus")

```

```{r rdp_taxonomy}
# Not run:
# taxa.rdp = assignTaxonomy(seqtab, './RDP/rdp_train_set_16.fa.gz', multithread=T)
# taxa.rdp = addSpecies(taxa.rdp, "./RDP/rdp_species_assignment_16.fa.gz")
# gc()
# colSums(!is.na(taxa.rdp))
```

```{r UNITE_taxonomy}

# taxa.UNITE = assignTaxonomy(seqtab.nochim, './UNITE/sh_general_release_dynamic_s_02.02.2019.fasta', multithread = TRUE, tryRC = TRUE)
# gc()
# colSums(!is.na(taxa.UNITE))
# Kingdom  Phylum   Class   Order  Family   Genus Species 
#    1917       2       1       1       1       1       1 

# ---> There are not many/any fungal reads in this dataset.

```


```{r idTaxa}
# TODO check out this idTaxa strategy on the new tutorial:
# https://benjjneb.github.io/dada2/tutorial.html
# Alternatives: The recently developed IdTaxa taxonomic classification method is also available via 
# the DECIPHER Bioconductor package. The paper introducing the IDTAXA algorithm reports classification 
# performance that is better than the long-time standard set by the naive Bayesian classifier. 
# Here we include a code block that allows you to use IdTaxa as a drop-in replacement for assignTaxonomy
# (and it’s faster as well!). Trained classifiers are available from http://DECIPHER.codes/Downloads.html.
# Download the SILVA SSU r132 (modified) file to follow along.

# library(DECIPHER); packageVersion("DECIPHER")

# dna <- DNAStringSet(getSequences(seqtab.nochim)) # Create a DNAStringSet from the ASVs
# load("./IDTAXA/SILVA_SSU_r138_2019.RData") # CHANGE TO THE PATH OF YOUR TRAINING SET
# #ids <- IdTaxa(dna, trainingSet, strand="top", processors=NULL, verbose=FALSE, threshold=0) # use all processors
# ids50 <- IdTaxa(dna, trainingSet, strand="top", processors=NULL, verbose=FALSE, threshold=50) # use all processors
# ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species") # ranks of interest
# # Convert the output object of class "Taxa" to a matrix analogous to the output from assignTaxonomy
# taxid <- t(sapply(ids50, function(x) {
#         m <- match(ranks, x$rank)
#         taxa <- x$taxon[m]
#         taxa[startsWith(taxa, "unclassified_")] <- NA
#         taxa
# }))
# #colnames(taxid) <- ranks; # c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
# # set caps so they match DADA2 
# colnames(taxid) = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species") output 
# rownames(taxid) <- getSequences(seqtab.nochim)
# saveRDS(taxid, file="IDTAXA_taxid.RDS")

# colSums(!is.na(taxid))

#  Domain  Phylum   Class   Order  Family   Genus Species 
#    1941    1695    1630    1500    1249     685       0 


```

```{r Compare_Zymo_Mock}
### There were no zymo/mock community samples in this experiment:
# zymo.genus = c('Pseudomonas', 'Escherichia/Shigella', 'Salmonella', 'Lactobacillus', 'Enterococcus', 'Staphylococcus', 'Listeria',
#     'Bacillus', 'Saccharomyces', 'Cryptococcus')

# unqs.mock = seqtab.nochim[grep("Zymo", rownames(seqtab.nochim)), ]
# unqs.mock = unqs.mock[,colSums(unqs.mock) > 0]  # drop ASVs with 0 count in mock
# unqs.mock = unqs.mock[,order(colSums(unqs.mock), decreasing=T)]  # order ASVs by count
# mock.countTable = data.frame(taxa.silva.nochim[colnames(unqs.mock), ], stringsAsFactors=F)
# mock.countTable$InZymo =  mock.countTable$Genus %in% zymo.genus
# mock.countTable = data.frame(mock.countTable, t(unqs.mock), stringsAsFactors=F)
# write.table(mock.countTable, file="Mock_count_table.tsv", row.names=F, col.names=T, sep='\t')

# mock.ASVs = DNAStringSet(colnames(unqs.mock))
# names(mock.ASVs) = paste0("ASV.", colSums(unqs.mock), "_", 
# apply(mock.countTable[,c('Class', "Order","Family", "Genus", "Species")], 1, paste, collapse='.'))
# writeXStringSet(mock.ASVs, file="mock_ASVs.fasta", format="fasta")
 
```

```{r countables}

## Create Amplicon Sequence Variant (ASV) table with counts:
CountTable.silva = data.frame(taxa.silva.nochim[colnames(seqtab.nochim), ], t(seqtab.nochim))
CountTable.silva = CountTable.silva[order(rowSums(t(seqtab.nochim)), decreasing=T), ]
write.table(data.frame(Sequence=rownames(CountTable.silva), CountTable.silva), row.names=F, file="AllSamples_AmpliconSequenceVariant_count_SILVA.tsv", sep='\t')

# CountTable.rdp = data.frame(taxa.rdp[colnames(seqtab), ], t(seqtab))
# CountTable.rdp = CountTable.rdp[order(rowSums(t(seqtab)), decreasing=T), ]
# write.table(data.frame(Sequence=rownames(CountTable.rdp), CountTable.rdp), row.names=F, file="AllSamples_AmpliconSequenceVariant_count_RDP.tsv", sep='\t')

```


```{r load_metadata}
# Load metadata for phyloseq object, re-order to match seqtab:
#mdata = read.table("../../SampleMetadata-compiled.tsv", header=T, as.is=T, sep='\t')
mdata = read.table("../../Master_DNA_SampleList_yr2-Modified.tsv", header=T, as.is=T, sep='\t')
#mdata2 = mdata[match(rownames(seqtab.nochim), mdata$SampleID), ]

### DOUBLE check that there are no missing samples in sample sheet or files:
setdiff(rownames(seqtab.nochim), mdata$SampleID) 

length(setdiff(rownames(seqtab.nochim), mdata$SampleID)) == 0

mdata2 = mdata[match(rownames(seqtab.nochim), mdata$SampleID), ]
rownames(mdata2) = mdata2$SampleID


kable(mdata2) %>%
  kable_styling("striped", full_width = F) %>%
  row_spec(0, angle = 0)

```

## Make a phylogenetic tree from all of the ASV sequences
```{r maketree}
# https://f1000research.com/articles/5-1492

library("DECIPHER")
library("phangorn")

ASVs.nochim = DNAStringSet(colnames(seqtab.nochim))
names(ASVs.nochim) = paste0("ASV", 1:ncol(seqtab.nochim))

alignment = AlignSeqs(DNAStringSet(ASVs.nochim), anchor=NA, processors=30)

phang.align <- phyDat(as(alignment, "matrix"), type="DNA") # turn into phyDat format
dm <- dist.ml(phang.align) # 
treeNJ <- NJ(dm) # Note, tip order != sequence order
fit = pml(treeNJ, data=phang.align)  # compute sthe likelihood of a phylogenetic tree given a sequence alignment and a model

## negative edges length changed to 0!
fitGTR <- update(fit, k=4, inv=0.2)
# optim.pml optimizes the different model parameters, this essentially searches for a better tree using a bunch of
# stochasitc rearrangement strategies.

### TODO
fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE,
                      rearrangement = "stochastic", control = pml.control(trace = 0))
#detach("package:phangorn", unload=TRUE)
save(fitGTR, file="fitGTR.RData")
```


```
```{r make_phyloseq_objects}
# Make Phyloseq objects
library(phyloseq)
library(Biostrings)

load("fitGTR.RData")
load("seqtab.nochim.RData")
load('taxa.silva.nochim.RData')

# create a Phyloseq object with Silva annotation:
ASVs.nochim = DNAStringSet(colnames(seqtab.nochim))
names(ASVs.nochim) = paste0("ASV", 1:ncol(seqtab.nochim))

tmp.seqtab = seqtab.nochim
colnames(tmp.seqtab) = names(ASVs.nochim)
tmp.taxa = taxa.silva.nochim
rownames(tmp.taxa) = names(ASVs.nochim)
ps.silva.nochim = phyloseq(otu_table(tmp.seqtab, taxa_are_rows=FALSE),
             sample_data(mdata2),
             tax_table(tmp.taxa),
             refseq(ASVs.nochim),
             phy_tree(fitGTR$tree))
save(ps.silva.nochim, file="phyloseq_nochim_silva.RData")

# Create a Phyloseq object with IDTAXA annotation:
# tmp.seqtab = seqtab.nochim
# colnames(tmp.seqtab) = names(ASVs.nochim)
# tmp.taxa = taxid
# rownames(tmp.taxa) = names(ASVs.nochim)

# ps.idtaxa.nochim = phyloseq(otu_table(tmp.seqtab, taxa_are_rows=FALSE),
#              sample_data(mdata2),
#              tax_table(tmp.taxa),
#              refseq(ASVs.nochim),
#              phy_tree(fitGTR$tree))
# save(ps.idtaxa.nochim, file="phyloseq_nochim_idtaxa.RData")

``` 

```{r}
# track <- cbind(out, final=rowSums(seqtab[,w]),perc=rowSums(seqtab[,w])/out$reads.in)
# rownames(track) <- sample.names
# write.table(track, file="track.tsv", row.names=F, sep='\t')


kable(track) %>%
  kable_styling("striped", full_width = F) %>%
  row_spec(0, angle = 0)

```
